---
title: "Warming tolerance of fish is remarkably consistent across life stages"
subtitle: "Data processing and analyses"
author: "Cowan et al. 2026"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
    embed-resources: true
    theme: cosmo
    fig-width: 8
    fig-height: 6
execute:
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| include: false

# Run every chunk from the project root, wherever the .qmd itself sits.
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(dev = "png", dpi = 500)
```

```{r}
#| label: packages

library(readxl)     # read the .xlsx workbook
library(dplyr)      # data wrangling
library(tidyr)      # pivot_wider / reshaping
library(ggplot2)    # figures
library(patchwork)  # compose the per-species + summary panel figure
library(brms)       # Bayesian mixed model
library(emmeans)    # estimated marginal means & contrasts
library(posterior)  # convergence diagnostics (summarise_draws)
library(coda)       # as.mcmc() to extract posterior draws of emmeans
library(multcompView)  # compact letter display for Figure 1

# Reproducibility & Stan backend --------------------------------------------
set.seed(123)
options(brms.backend = "cmdstanr")   # cmdstan is installed; faster & lighter

# Data source
xlsx <- "data/Life_stages_project_all_data.xlsx"

# The four life stages, in developmental order -------------------------------
STAGE_LEVELS <- c("Embryo", "Larva", "Immature", "Spawner")

# Colour palette for each species
species_cols <- c(Atlantic_salmon = "#F4A261", Brook_trout   = "#6A994E",
                  Sand_goby       = "#7570B3", Two_spot_goby = "#E9C46A",
                  Stickleback     = "#0B4F6C", Black_goby    = "#A26769",
                  Pumpkinseed     = "#C06C84", Zebrafish     = "#43AA8B")

# Parse numbers that may use a decimal comma (European locale) or the literal
# string "NA". Bad values become NA.
num_eu <- function(x) suppressWarnings(as.numeric(gsub(",", ".", as.character(x))))

# Trim the unicode spaces base trimws() leaves in place. Some stage labels
# ("Juvenile ") end in a non-breaking space, which breaks the recode() calls.
norm_ws <- function(x) {
  weird <- intToUtf8(c(0x00A0, 0x2007, 0x200B, 0x202F, 0x3000), multiple = TRUE)
  x <- gsub(paste0("[", paste(weird, collapse = ""), "]"), " ", as.character(x))
  trimws(x)
}
```

# Data import and cleaning

Each block reads the worksheet for one species. We excluded embryos that hatched
during the trial, because hatching removes the CTmax endpoint. We also excluded
missing CTmax values and records flagged with other problems. We standardised
the life-stage columns to four levels: embryo, larva, immature and spawner.

## Zebrafish

Load and filter data for zebrafish (*Danio rerio*).

```{r}
#| label: clean-zebrafish

zebrafish <- read_excel(xlsx, "Zebrafish") %>%
  filter(is.na(Hatched_during_trial) | Hatched_during_trial != "Y") %>%
  filter(!is.na(CTmax)) %>%
  mutate(
    Species       = "Zebrafish",
    LifeStage     = recode(norm_ws(Lifestage), "Larvae" = "Larva", "Juvenile" = "Immature"),
    CTmax         = as.numeric(CTmax),
    UniqueTrialID = paste("Zebrafish", LifeStage, Trial_no),
    Ramping_rate  = 18
  ) %>%
  select(Species, LifeStage, CTmax, UniqueTrialID, Ramping_rate)
```

## Two-spot goby

Load and filter data for two-spot goby (*Gobiusculus flavescens*).

```{r}
#| label: clean-twospot

two_spot <- read_excel(xlsx, "Two_spot") %>%
  mutate(CTmax = num_eu(CTmax)) %>%
  filter(!is.na(CTmax)) %>%
  mutate(
    Species       = "Two_spot_goby",
    LifeStage     = recode(norm_ws(Lifestage),
                           "Larvae" = "Larva", "Juvenile" = "Immature", "Adult" = "Spawner"),
    UniqueTrialID = paste("Two_spot_goby", LifeStage, Trial_nr.),
    Ramping_rate  = 18
  ) %>%
  select(Species, LifeStage, CTmax, UniqueTrialID, Ramping_rate)
```

## Sand goby

Load and filter data for sand goby (*Pomatoschistus minutus*).

```{r}
#| label: clean-sandgoby

sand_goby <- read_excel(xlsx, "Sand_goby") %>%
  mutate(
    Species       = "Sand_goby",
    LifeStage     = recode(norm_ws(LifeStage), "Larvae" = "Larva", "Non-spawner" = "Immature"),
    CTmax         = as.numeric(CTmax),
    UniqueTrialID = paste("Sand_goby", LifeStage, Trial_no),
    Ramping_rate  = 18
  ) %>%
  filter(!is.na(CTmax)) %>%
  select(Species, LifeStage, CTmax, UniqueTrialID, Ramping_rate)
```

## Black goby

Load and filter data for black goby (*Gobius niger*). We also excluded one
faulty record (trial 9, individual 1) and seven observations for larvae in which
the assay was compromised.

```{r}
#| label: clean-blackgoby

bg <- read_excel(xlsx, "Black_goby") %>% mutate(LifeStage = norm_ws(LifeStage))

bg_embryo <- bg %>%
  filter(LifeStage == "Embryo") %>%
  filter(is.na(Hatched) | Hatched != "Yes") %>%
  filter(!(Trial_no == 9 & Individual == 1)) %>%
  filter(!is.na(CTmax))

bg_larva <- bg %>%
  filter(LifeStage == "Larvae") %>%
  filter(!(Individual %in% c(1, 20, 37, 38, 39, 40, 47)))

bg_adult <- bg %>%
  filter(LifeStage %in% c("Adult", "Spawner"))

black_goby <- bind_rows(bg_embryo, bg_larva, bg_adult) %>%
  mutate(
    Species    = "Black_goby",
    LifeStage  = recode(LifeStage, "Larvae" = "Larva", "Adult" = "Immature"),
    trial_stg  = if_else(LifeStage %in% c("Immature", "Spawner"), "Adult", LifeStage),
    UniqueTrialID = paste("Black_goby", trial_stg, Trial_no),
    CTmax      = as.numeric(CTmax),
    Ramping_rate = 18
  ) %>%
  filter(!is.na(CTmax)) %>%
  select(Species, LifeStage, CTmax, UniqueTrialID, Ramping_rate)
```

## Three-spined stickleback

Load and filter data for three-spined stickleback (*Gasterosteus aculeatus*).

```{r}
#| label: clean-stickleback

stickleback <- read_excel(xlsx, "Stickleback") %>%
  filter(is.na(Hatched) | Hatched != "Yes") %>%
  mutate(CTmax = as.numeric(CTmax)) %>%
  filter(!is.na(CTmax)) %>%
  mutate(
    Species       = "Stickleback",
    LifeStage     = recode(norm_ws(LifeStage), "Larvae" = "Larva"),
    UniqueTrialID = paste("Stickleback", LifeStage, Trial_no),
    Ramping_rate  = 18
  ) %>%
  select(Species, LifeStage, CTmax, UniqueTrialID, Ramping_rate)
```

## Atlantic salmon

Load and filter data for Atlantic salmon (*Salmo salar*). We treated alevins as
larvae and post-smolts as immatures.

```{r}
#| label: clean-salmon

salmon <- read_excel(xlsx, "Atlantic_salmon") %>%
  filter(is.na(Hatched) | Hatched != "Y") %>%
  mutate(CTmax = num_eu(CTmax)) %>%
  filter(!is.na(CTmax)) %>%
  mutate(
    Species       = "Atlantic_salmon",
    LifeStage     = recode(norm_ws(Life_stage), "Alevin" = "Larva", "PostSmolt" = "Immature"),
    UniqueTrialID = paste("Atlantic_salmon", LifeStage, Trial_no),
    Ramping_rate  = `Ramping_rate_C/h`
  ) %>%
  filter(LifeStage %in% STAGE_LEVELS) %>%
  select(Species, LifeStage, CTmax, UniqueTrialID, Ramping_rate)
```

## Brook trout

Load and filter data for brook trout (*Salvelinus fontinalis*). The larval stage
was not assayed for this species. Of the 102 fish first recorded as spawners, 55
could not be confidently assigned as spawners and were thus excluded.

```{r}
#| label: clean-brooktrout

brook_trout <- read_excel(xlsx, "Brook_trout") %>%
  filter(!is.na(CTmax)) %>%
  mutate(
    Species       = "Brook_trout",
    LifeStage     = recode(norm_ws(life_stage), "Juvenile" = "Immature"),
    CTmax         = as.numeric(CTmax),
    UniqueTrialID = paste("Brook_trout", LifeStage, trial),
    Ramping_rate  = 18
  ) %>%
  filter(LifeStage %in% STAGE_LEVELS) %>%
  select(Species, LifeStage, CTmax, UniqueTrialID, Ramping_rate)
```

## Pumpkinseed sunfish

Load and filter data for pumpkinseed sunfish (*Lepomis gibbosus*).

```{r}
#| label: clean-pumpkinseed

pumpkinseed <- read_excel(xlsx, "Pumpkinseed") %>%
  mutate(Stage = norm_ws(Stage)) %>%
  mutate(
    Species   = "Pumpkinseed",
    LifeStage = recode(Stage,
                       "Reproductive Adult" = "Spawner", "Adult" = "Immature", "Larvae" = "Larva"),
    CTmax     = if_else(LifeStage %in% c("Immature", "Spawner"),
                        as.numeric(Ctmax_Han), as.numeric(Ctmax_FireSting)),
    UniqueTrialID = paste("Pumpkinseed", LifeStage, Trial),
    Ramping_rate  = 18
  ) %>%
  filter(!is.na(CTmax)) %>%
  select(Species, LifeStage, CTmax, UniqueTrialID, Ramping_rate)
```

## Combine all species

CTmax varies with ramping rate, so we only analysed measurements taken using a
consistent ramping rate across life stages (18 °C h⁻¹).

```{r}
#| label: combine

AllSpecies <- bind_rows(
  zebrafish, two_spot, sand_goby, black_goby,
  stickleback, salmon, brook_trout, pumpkinseed
) %>%
  mutate(
    LifeStage     = factor(LifeStage, levels = STAGE_LEVELS),
    Species       = factor(Species),
    UniqueTrialID = factor(UniqueTrialID)
  )

# a stage label outside STAGE_LEVELS would silently become NA above
stopifnot(!anyNA(AllSpecies$LifeStage))
# every retained measurement used the same ramping rate
stopifnot(all(AllSpecies$Ramping_rate == 18))

write.csv(AllSpecies, "output/AllSpecies_clean.csv", row.names = FALSE)
```

## Sample sizes

Total sample sizes after data curation.

```{r}
#| label: total-sample-size

total_tbl <- data.frame(
  Individuals = nrow(AllSpecies),
  Trials      = nlevels(droplevels(AllSpecies$UniqueTrialID)),
  Species     = nlevels(droplevels(AllSpecies$Species))
)
knitr::kable(total_tbl, caption = "Total sample sizes after data curation.")
```

Individuals per species and life stage.

```{r}
#| label: sample-sizes

sp_labels <- c(
  Atlantic_salmon = "Atlantic salmon (<i>Salmo salar</i>)",
  Black_goby      = "Black goby (<i>Gobius niger</i>)",
  Brook_trout     = "Brook trout (<i>Salvelinus fontinalis</i>)",
  Pumpkinseed     = "Pumpkinseed sunfish (<i>Lepomis gibbosus</i>)",
  Sand_goby       = "Sand goby (<i>Pomatoschistus minutus</i>)",
  Stickleback     = "Three-spined stickleback (<i>Gasterosteus aculeatus</i>)",
  Two_spot_goby   = "Two-spot goby (<i>Gobiusculus flavescens</i>)",
  Zebrafish       = "Zebrafish (<i>Danio rerio</i>)"
)

n_wide <- AllSpecies %>%
  count(Species, LifeStage) %>%
  pivot_wider(names_from = LifeStage, values_from = n, values_fill = 0) %>%
  arrange(Species) %>%
  select(Species, all_of(STAGE_LEVELS))
n_wide$Total <- rowSums(n_wide[, STAGE_LEVELS])

write.csv(n_wide, "output/sample_sizes.csv", row.names = FALSE)

col_totals <- colSums(n_wide[, c(STAGE_LEVELS, "Total")])
disp <- n_wide %>% mutate(Species = sp_labels[as.character(Species)])
disp <- rbind(disp,
              data.frame(Species = "<b>Total</b>", as.list(col_totals),
                         check.names = FALSE))

knitr::kable(disp, escape = FALSE, align = c("l", rep("r", 5)),
             caption = "Sample sizes per species and life stage.")
```

# Bayesian mixed model (brms)

We fitted a Bayesian linear mixed model to model variation in upper thermal
limits across life stages using the brms package (Bürkner 2017; version 2.23.0).
We fitted life stage as a fixed effect without an intercept, so that mean upper
thermal limits were estimated for each stage rather than as a difference from a
reference stage. We used species identity as a random effect with uncorrelated
life-stage slopes. Because individuals tested within the same trial are not
independent, we also added trial identity as a random intercept. The model
structure was as follows:

$$\texttt{CTmax} \sim 0 + \texttt{LifeStage} + (0 + \texttt{LifeStage}\ ||\ \texttt{Species}) + (1\ |\ \texttt{UniqueTrialID})$$

During data exploration, we also noticed that CTmax varied far more among
embryos than later life stages (i.e. heteroscedasticity). We therefore let the
residual standard deviation differ among stages (`sigma ~ 0 + LifeStage`).

We used weakly informative priors: `Normal(30, 10)` on the life-stage means,
`Normal(0, 2)` on the log-scale residual standard deviations, and half
`Student-t(3, 0, 5)` on the random-effect standard deviations. We ran four
chains of 8,000 iterations, with adapt_delta set to 0.95 and maximum tree depth
to 12, and checked convergence using R-hat, effective sample sizes, divergent
transitions and E-BFMI.

## Fit

```{r}
#| label: fit-model-random
#| output: false

form <- bf(
  CTmax ~ 0 + LifeStage + (0 + LifeStage || Species) + (1 | UniqueTrialID),
  sigma ~ 0 + LifeStage
)
priors <- c(
  prior(normal(30, 10),     class = "b"),                  # mean CTmax per stage
  prior(normal(0, 2),       class = "b", dpar = "sigma"),  # log residual SD per stage
  prior(student_t(3, 0, 5), class = "sd")                  # random-effect SDs
)
brms_model <- brm(
  formula = form, family = gaussian(), data = AllSpecies,
  prior = priors, chains = 4, cores = 4, iter = 8000, seed = 123,
  control = list(adapt_delta = 0.95, max_treedepth = 12),
  silent  = 2, refresh = 0,
  file        = "output/brms_model",
  file_refit  = "on_change"
)
```

```{r}
#| label: model-summary-random
summary(brms_model)
```

## Diagnostics

```{r}
#| label: diag-random

# R-hat, ESS, divergences, treedepth and E-BFMI against their usual thresholds.
sampler_diagnostics <- function(model, max_td) {
  np <- brms::nuts_params(model)
  nd <- brms::ndraws(model)
  conv <- posterior::summarise_draws(posterior::as_draws_df(model),
                                     "rhat", "ess_bulk", "ess_tail")
  ebfmi <- np |>
    dplyr::filter(Parameter == "energy__") |>
    dplyr::group_by(Chain) |>
    dplyr::summarise(e = sum(diff(Value)^2) / sum((Value - mean(Value))^2),
                     .groups = "drop")
  ndiv <- sum(np$Value[np$Parameter == "divergent__"])
  ntd  <- sum(np$Value[np$Parameter == "treedepth__"] == max_td)
  mr   <- max(conv$rhat, na.rm = TRUE)
  eb   <- min(conv$ess_bulk, na.rm = TRUE)
  et   <- min(conv$ess_tail, na.rm = TRUE)
  me   <- min(ebfmi$e)
  data.frame(
    Diagnostic = c("Max split-R-hat", "Min bulk-ESS", "Min tail-ESS",
                   "Divergent transitions", "Max-treedepth hits", "Min E-BFMI"),
    Value = c(round(mr, 4), round(eb), round(et), ndiv, ntd, round(me, 3)),
    Threshold = c("< 1.01", "> 400", "> 400",
                  paste0("< 0.5% of ", nd), paste0("< 0.5% of ", nd), "> 0.2"),
    Status = c(ifelse(mr < 1.01, "OK", "CHECK"),
               ifelse(eb > 400, "OK", "CHECK"),
               ifelse(et > 400, "OK", "CHECK"),
               ifelse(ndiv / nd < 0.005, "OK", "CHECK"),
               ifelse(ntd / nd < 0.005, "OK", "CHECK"),
               ifelse(me > 0.2, "OK", "CHECK"))
  )
}

diagnostics <- sampler_diagnostics(brms_model, max_td = 12)
write.csv(diagnostics, "output/model_diagnostics.csv", row.names = FALSE)
knitr::kable(diagnostics,
             caption = "Sampler convergence diagnostics (all parameters).")
```

```{r}
#| label: diag-random-plots
#| fig-width: 9
#| fig-height: 6

brms::mcmc_plot(brms_model, type = "trace", variable = "^b_", regex = TRUE)
pp_check(brms_model, ndraws = 100)
```

## Overall life-stage means

Here, we present the predicted CTmax for each life stage, across species. We
used emmeans to generate posterior medians and 95% highest posterior density
intervals from the fitted model.

```{r}
#| label: emmeans-overall-means

emm_overall <- emmeans(brms_model, ~ LifeStage, re_formula = NA)
overall_means_tab <- as.data.frame(
  summary(emm_overall, point.est = median, level = 0.95)
)
write.csv(overall_means_tab, "output/overall_lifestage_means.csv", row.names = FALSE)
knitr::kable(overall_means_tab, digits = 2,
             caption = "Overall life-stage marginal means (posterior median, 95% HPD interval).")
```

## Overall life-stage contrasts

Here, we present contrasts between life stages, estimated using emmeans from the
fitted model.

```{r}
#| label: emmeans-overall-random

overall_tab <- as.data.frame(
  summary(pairs(emm_overall, adjust = "none"), point.est = median, level = 0.95)
)
write.csv(overall_tab, "output/overall_lifestage_contrasts.csv",
          row.names = FALSE)
knitr::kable(overall_tab, digits = 3,
             caption = "Overall pairwise life-stage contrasts (posterior median, 95% HPD interval).")
```

## Species-specific life-stage contrasts

Here, we present species-specific contrasts between life stages. We estimated
these life-stage differences using emmeans and present posterior medians and 95%
highest posterior density intervals. Note that we did not measure the CTmax of
brook trout larvae, but the model returns a prediction for this life stage
through Bayesian data augmentation.

```{r}
#| label: emmeans-species-random

emm_species <- emmeans(brms_model, ~ LifeStage | Species,
                              re_formula = ~ (0 + LifeStage || Species))
species_tab <- as.data.frame(
  summary(pairs(emm_species, adjust = "none"), point.est = median, level = 0.95)
) %>%
  mutate(Species = gsub("_", " ", as.character(Species)))
write.csv(species_tab, "output/species_lifestage_contrasts.csv",
          row.names = FALSE)
knitr::kable(species_tab, digits = 3,
             caption = "Species-specific pairwise life-stage contrasts (posterior median, 95% HPD interval).")
```

# Figures

## Figure 1: CTmax across life stages in all eight species

```{r}
#| label: fig1-data

# Species ordered by acclimation temperature, which is also the panel order.
fig1_levels <- c("Atlantic_salmon", "Brook_trout", "Sand_goby", "Two_spot_goby",
                 "Stickleback", "Black_goby", "Pumpkinseed", "Zebrafish")
fig1_labs <- c(
  Atlantic_salmon = "Atlantic salmon (acclimated to 6 °C)",
  Brook_trout     = "Brook trout (acclimated to 8 °C)",
  Sand_goby       = "Sand goby (acclimated to 11 °C)",
  Two_spot_goby   = "Two-spot goby (acclimated to 11 °C)",
  Stickleback     = "Three-spined stickleback (acclimated to 18 °C)",
  Black_goby      = "Black goby (acclimated to 18 °C)",
  Pumpkinseed     = "Pumpkinseed sunfish (acclimated to 21 °C)",
  Zebrafish       = "Zebrafish (acclimated to 28 °C)"
)

fig1_dat <- AllSpecies %>%
  mutate(Species = factor(as.character(Species), levels = fig1_levels),
         # translucent fill baked in, so the outline can stay fully opaque
         pt_fill = grDevices::adjustcolor(
           unname(species_cols[as.character(Species)]), alpha.f = 0.4))

# mean and its 95% CI across the individuals of a species x stage cell
fig1_summ <- fig1_dat %>%
  group_by(Species, LifeStage) %>%
  summarise(n = n(), mean_CTmax = mean(CTmax), se = sd(CTmax) / sqrt(n()),
            .groups = "drop") %>%
  mutate(lo = mean_CTmax - qt(0.975, n - 1) * se,
         hi = mean_CTmax + qt(0.975, n - 1) * se)

# Letters from the species-specific contrasts: two stages differ when the 95% HPD
# interval of their contrast excludes zero. Brook trout larvae were not assayed,
# so contrasts involving them are left out.
fig1_letters <- species_tab %>%
  mutate(Species  = gsub(" ", "_", Species),
         contrast = gsub(" - ", "-", as.character(contrast)),
         differ   = lower.HPD > 0 | upper.HPD < 0) %>%
  filter(!(Species == "Brook_trout" & grepl("Larva", contrast))) %>%
  group_by(Species) %>%
  group_modify(~ {
    lt <- multcompLetters(setNames(.x$differ, .x$contrast))$Letters
    data.frame(LifeStage = names(lt), letter = unname(lt))
  }) %>%
  ungroup() %>%
  mutate(Species   = factor(Species, levels = fig1_levels),
         LifeStage = factor(LifeStage, levels = STAGE_LEVELS)) %>%
  left_join(fig1_dat %>%
              group_by(Species, LifeStage) %>%
              summarise(ymax = max(CTmax), .groups = "drop"),
            by = c("Species", "LifeStage"))
```

```{r}
#| label: fig-allspecies
#| fig-width: 11
#| fig-height: 12
#| fig-cap: "Thermal tolerance is relatively conserved across life stages (embryo, larva, immature adult, spawner). Acute upper thermal tolerance (CTmax) across four life stages in eight fish species (note that there is no larval data for brook trout). Each coloured point represents one individual, jittered horizontally; black points and whiskers show the mean CTmax ± 95% confidence interval for each life stage. Letters compare life stages within each species: stages that share a letter did not differ, because the 95% highest posterior density interval of their contrast in the Bayesian mixed model included zero."

p_allspecies <- ggplot(fig1_dat, aes(LifeStage, CTmax, colour = Species)) +
  # translucent fill with an opaque outline of the same colour
  geom_jitter(aes(fill = pt_fill), shape = 21, width = 0.25, height = 0,
              size = 3.5, stroke = 0.4) +
  scale_fill_identity(guide = "none") +
  geom_errorbar(data = fig1_summ, aes(LifeStage, ymin = lo, ymax = hi),
                width = 0.25, linewidth = 0.6, colour = "black", inherit.aes = FALSE) +
  geom_point(data = fig1_summ, aes(LifeStage, mean_CTmax),
             colour = "black", size = 1.8, inherit.aes = FALSE) +
  # letters sit just above the highest point of each life stage
  geom_text(data = fig1_letters, aes(LifeStage, pmin(ymax + 2, 46), label = letter),
            vjust = 0, size = 4, inherit.aes = FALSE) +
  facet_wrap(~ Species, nrow = 4, labeller = labeller(Species = fig1_labs)) +
  scale_colour_manual(values = species_cols, guide = "none") +
  scale_x_discrete(limits = STAGE_LEVELS, drop = FALSE) +
  scale_y_continuous(limits = c(13, 47), breaks = seq(10, 45, 5)) +
  labs(x = NULL, y = "Upper thermal limit (°C)") +
  theme_classic(base_size = 12) +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.7),
        axis.line = element_blank(), panel.grid = element_blank(),
        panel.background = element_rect(fill = "white", colour = NA),
        strip.background = element_blank(), strip.text = element_text(size = 10.5),
        axis.text = element_text(colour = "black"),
        # larger y title, pushed off the tick labels
        axis.title.y = element_text(size = 16, margin = margin(r = 12)))
p_allspecies

ggsave("fig/Figure1_CTmax_life_stages.jpg", p_allspecies, width = 11, height = 12, dpi = 300)
ggsave("fig/Figure1_CTmax_life_stages.pdf", p_allspecies, width = 11, height = 12)
```

## Figure 2: comparison with Dahlke et al. (2020)

Here, we provide a comparison of our estimates to a comparative analysis from
Dahlke et al. (2020). Our estimates were all produced using a consistent
protocol across all eight species and four life stages. Dahlke et al. (2020)
compiled upper thermal limits for 764 fish species, the broadest assessment of
differences in heat tolerance between fish life stages to date. We compare the
two datasets across the six species they share (Atlantic salmon, brook trout,
sand goby, three-spined stickleback, black goby and zebrafish).

From our data, we took the posterior median and 95% credible interval
from our fitted Bayesian linear mixed model. Because we did not measure the
CTmax of brook trout larvae, we left this value empty. Dahlke et al. (2020)
reported one upper thermal limit estimate per species and life stage, either
measured experimentally or imputed using phylogenetic data imputation
(phylopars).

The mean CTmax spanned 14.8 °C among the six species, from 26.2 °C in Atlantic
salmon to 41.0 °C in zebrafish. Therefore, in panel (b) we expressed the CTmax
of each life stage relative to its own immature stage before averaging, to place
all six species on a similar scale. In panel (c), we report the difference
between the warmest and coolest stage within each species. In both panels, we
report a mean and 95% confidence interval across the six species.

```{r}
#| label: data-species-cmp

dahlke_stage <- c(a_Spawner = "Spawner", b_Embryo = "Embryo",
                  c_Larvae = "Larva",   d_Adult  = "Immature")

sp6 <- c("Atlantic_salmon", "Brook_trout", "Sand_goby",
         "Stickleback", "Black_goby", "Zebrafish")
species_labs <- c(Atlantic_salmon = "Atlantic salmon", Brook_trout = "Brook trout",
                  Sand_goby = "Sand goby", Stickleback = "Three-spined stickleback",
                  Black_goby = "Black goby", Zebrafish = "Zebrafish")
sci2int <- c("Salmo salar" = "Atlantic_salmon", "Salvelinus fontinalis" = "Brook_trout",
             "Pomatoschistus minutus" = "Sand_goby", "Gasterosteus aculeatus" = "Stickleback",
             "Gobius niger" = "Black_goby", "Danio rerio" = "Zebrafish")

# our per species x stage means
this_sp <- AllSpecies |>
  filter(Species %in% sp6) |>
  group_by(Species, LifeStage) |>
  summarise(se = sd(CTmax) / sqrt(n()), CTmax = mean(CTmax), .groups = "drop") |>
  mutate(Dataset = "This study", Imputed = FALSE)

# Dahlke: one value per species x stage, flagged as measured or imputed
dahlke_sp <- read_excel(
    "data/Experimental_and_imputed_tolerance_data_Dahlke_et_al_2020.xlsx", "Tmax") |>
  filter(SpeciesFishBase %in% names(sci2int)) |>
  transmute(Species = unname(sci2int[SpeciesFishBase]),
            LifeStage = factor(unname(dahlke_stage[Lifestage]), levels = STAGE_LEVELS),
            CTmax = `Tmax (°C)`, se = NA_real_,
            Dataset = "Dahlke", Imputed = Parameter == "Imputed")

# model estimates and 95% credible intervals per species x stage
emm_sp_grid <- emmeans(brms_model, ~ LifeStage * Species,
                       re_formula = ~ (0 + LifeStage || Species))
mc_sp <- as.matrix(as.mcmc(emm_sp_grid))     # one column per cell, in grid order
sm_sp <- as.data.frame(summary(emm_sp_grid))
stopifnot(ncol(mc_sp) == nrow(sm_sp))

model_sp <- data.frame(
  Species   = as.character(sm_sp$Species),
  LifeStage = factor(as.character(sm_sp$LifeStage), levels = STAGE_LEVELS),
  est       = apply(mc_sp, 2, median),
  lo        = apply(mc_sp, 2, quantile, 0.025),
  hi        = apply(mc_sp, 2, quantile, 0.975),
  row.names = NULL) |>
  filter(Species %in% sp6) |>
  mutate(Species = factor(Species, levels = sp6))

sp_cmp <- bind_rows(this_sp, dahlke_sp) |>
  mutate(Species = factor(Species, levels = sp6),
         xnum = as.integer(LifeStage))

# One table for all three panels.
prov_lev <- c("This study", "Previously reported (measured)",
              "Previously reported (imputed)")
sp_panel <- sp_cmp |>
  left_join(model_sp, by = c("Species", "LifeStage")) |>
  mutate(CTmax = if_else(Dataset == "This study", est, CTmax),
         lo    = if_else(Dataset == "This study", lo, NA_real_),
         hi    = if_else(Dataset == "This study", hi, NA_real_),
         Provenance = factor(dplyr::case_when(
           Dataset == "This study" ~ "This study",
           Imputed                 ~ "Previously reported (imputed)",
           TRUE                    ~ "Previously reported (measured)"), levels = prov_lev),
         # every marker is outlined, so "open" means a white fill
         sp_col  = unname(species_cols[as.character(Species)]),
         pt_fill = if_else(Provenance == "Previously reported (imputed)", "white", sp_col),
         xint = as.integer(LifeStage))

# Each species relative to its own immature value. n_sp is 5 at Larva for this
# study, since brook trout has no larval data.
norm_sp <- sp_panel |>
  filter(!is.na(CTmax)) |>
  group_by(Dataset, Species) |>
  mutate(rel = CTmax - CTmax[LifeStage == "Immature"]) |>
  ungroup()

norm_avg <- norm_sp |>
  group_by(Dataset, LifeStage) |>
  summarise(n_sp = n(), se = sd(rel) / sqrt(n()), rel = mean(rel), .groups = "drop") |>
  mutate(lo = rel - qt(0.975, n_sp - 1) * se,
         hi = rel + qt(0.975, n_sp - 1) * se,
         dcol = ifelse(Dataset == "This study", "black", "grey45"),
         xnum = as.integer(LifeStage) + ifelse(Dataset == "This study", -0.05, 0.05))
```

```{r}
#| label: fig-species-panel
#| fig-width: 11
#| fig-height: 11
#| fig-cap: "Upper thermal limits across life stages in the six species studied here, compared with values previously reported by Dahlke et al. (2020). (a) Variation in upper thermal limits across life stages. Posterior medians from this study are represented by circles (with 95% credible intervals) while previously reported estimates are presented with coloured (measured) or empty (imputed) squares. (b) Relative variation in upper thermal limits between life stages. Each species CTmax is set to zero at its immature value. Mean estimates and 95% confidence intervals across species are represented for this study (circles) and for previously reported values (squares). Faint coloured lines represent estimates for each individual species (solid for this study, long-dashed for previously reported values). (c) Range of variation in upper thermal limits across life stages. Estimates represent the difference between the warmest and coolest life stage within each species, with the mean and its 95% confidence interval across species shown as the white point estimates and error bars."

stage_abbr <- c("Emb", "Lar", "Imm", "Spw")
col_this   <- "black"
col_dahlke <- "grey45"

# (a) -----------------------------------------------------------------------
p_facets <- ggplot(sp_panel, aes(xint, CTmax, colour = Species)) +
  geom_line(aes(linetype = Dataset, group = Dataset, linewidth = Dataset)) +
  scale_linewidth_manual(values = c("This study" = 1.3, "Dahlke" = 0.7),
                         guide = "none") +
  geom_linerange(data = filter(sp_panel, Dataset == "This study"),
                 aes(ymin = lo, ymax = hi), linewidth = 1.5) +
  geom_point(aes(shape = Provenance, fill = pt_fill),
             colour = "black", size = 2.6, stroke = 0.6) +
  facet_wrap(~ Species, nrow = 2, scales = "free_y",
             labeller = labeller(Species = function(x) species_labs[x])) +
  scale_colour_manual(values = species_cols, guide = "none") +
  scale_fill_identity(guide = "none") +
  scale_linetype_manual(values = c("This study" = "solid", "Dahlke" = "longdash"),
                        guide = "none") +
  scale_shape_manual(values = c("This study" = 21, "Previously reported (measured)" = 22,
                                "Previously reported (imputed)" = 22), name = NULL) +
  scale_x_continuous(breaks = 1:4, labels = stage_abbr, expand = expansion(add = 0.3)) +
  labs(x = NULL, y = "Upper thermal limit (°C)") +
  guides(shape = guide_legend(override.aes = list(
    colour = "black", fill = c("black", "black", "white"), linetype = 0))) +
  theme_classic(base_size = 12) +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.7),
        axis.line = element_blank(), panel.grid = element_blank(),
        panel.background = element_rect(fill = "white", colour = NA),
        strip.background = element_blank(), strip.text = element_text(size = 10.5),
        axis.text = element_text(colour = "black"),
        legend.position = c(0.006, 0.015), legend.justification = c(0, 0),
        legend.direction = "vertical",
        legend.background = element_rect(fill = "white", colour = NA),
        legend.key = element_blank(), legend.key.height = unit(0.38, "cm"),
        legend.spacing.y = unit(0.02, "cm"),
        legend.text = element_text(size = 8.5))

# (b) -----------------------------------------------------------------------
# Faint coloured lines are the individual species, behind the overall means.
lab_this   <- data.frame(x = 0.95, y = -21.2, lab = "This study")
lab_dahlke <- data.frame(x = 0.95, y = -23.4, lab = "Previously reported")
key_xy     <- data.frame(x = 0.85, y = c(-21.2, -23.4))

p_norm <- ggplot() +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey60", linewidth = 0.5) +
  geom_line(data = norm_sp, aes(xint, rel, colour = Species, linetype = Dataset,
            group = interaction(Species, Dataset)), alpha = 0.4, linewidth = 0.45) +
  geom_line(data = filter(norm_avg, Dataset == "Dahlke"), aes(xnum, rel),
            colour = col_dahlke, linewidth = 1.2, linetype = "longdash") +
  geom_line(data = filter(norm_avg, Dataset == "This study"), aes(xnum, rel),
            colour = col_this, linewidth = 1.75, linetype = "solid") +
  geom_rect(data = norm_avg, aes(xmin = xnum - 0.014, xmax = xnum + 0.014,
            ymin = lo, ymax = hi, fill = dcol)) +
  geom_point(data = filter(norm_avg, Dataset == "This study"), aes(xnum, rel),
             shape = 21, fill = "white", colour = "black", stroke = 0.8, size = 4) +
  geom_point(data = filter(norm_avg, Dataset == "Dahlke"), aes(xnum, rel),
             shape = 22, fill = "white", colour = "black", stroke = 0.8, size = 4) +
  # inline key: the same markers used for the two means
  geom_point(data = key_xy[1, ], aes(x, y), shape = 21, fill = "white",
             colour = "black", stroke = 0.8, size = 4) +
  geom_point(data = key_xy[2, ], aes(x, y), shape = 22, fill = "white",
             colour = "black", stroke = 0.8, size = 4) +
  geom_text(data = lab_this,   aes(x, y, label = lab), hjust = 0, size = 4.2,
            fontface = "bold", colour = col_this) +
  geom_text(data = lab_dahlke, aes(x, y, label = lab), hjust = 0, size = 4.2,
            fontface = "bold", colour = col_dahlke) +
  scale_colour_manual(values = species_cols, guide = "none") +
  scale_fill_identity(guide = "none") +
  scale_linetype_manual(values = c("This study" = "solid", "Dahlke" = "longdash"),
                        guide = "none") +
  scale_x_continuous(breaks = 1:4, labels = STAGE_LEVELS,
                     expand = expansion(add = c(0.3, 0.3))) +
  labs(x = NULL, y = "Upper thermal limit relative to immature (°C)") +
  theme_classic(base_size = 13) +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),
        axis.line = element_blank(), panel.grid = element_blank(),
        axis.text = element_text(colour = "black"))

# (c) -----------------------------------------------------------------------
# Warmest minus coolest stage within each species. Brook trout is compared over
# three stages rather than four.
sp_range <- sp_panel |>
  filter(!is.na(CTmax)) |>
  group_by(Dataset, Species) |>
  summarise(n_stage = n(), rng = max(CTmax) - min(CTmax), .groups = "drop") |>
  mutate(xnum    = ifelse(Dataset == "This study", 1, 1.6),
         xspread = xnum + (as.integer(Species) - 3.5) * 0.03)

sp_range_avg <- sp_range |>
  group_by(Dataset) |>
  summarise(n_sp = n(), se = sd(rng) / sqrt(n()), mean_rng = mean(rng), .groups = "drop") |>
  mutate(lo = mean_rng - qt(0.975, n_sp - 1) * se,
         hi = mean_rng + qt(0.975, n_sp - 1) * se,
         xnum = ifelse(Dataset == "This study", 1, 1.6))

# Half-width set so the bar prints the same thickness as those in panel (b).
p_range <- ggplot() +
  geom_point(data = sp_range, aes(xspread, rng, fill = Species, shape = Dataset),
             colour = "black", size = 3, stroke = 0.5) +
  geom_rect(data = filter(sp_range_avg, Dataset == "This study"),
            aes(xmin = xnum - 0.012, xmax = xnum + 0.012, ymin = lo, ymax = hi),
            fill = col_this, alpha = 0.95) +
  geom_rect(data = filter(sp_range_avg, Dataset == "Dahlke"),
            aes(xmin = xnum - 0.012, xmax = xnum + 0.012, ymin = lo, ymax = hi),
            fill = col_dahlke, alpha = 0.95) +
  geom_point(data = sp_range_avg, aes(xnum, mean_rng, shape = Dataset),
             fill = "white", colour = "black", size = 4.4, stroke = 0.9) +
  scale_fill_manual(values = species_cols, labels = species_labs, name = NULL) +
  scale_shape_manual(values = c("This study" = 21, "Dahlke" = 22), guide = "none") +
  scale_x_continuous(breaks = c(1, 1.6), labels = c("This study", "Previously reported"),
                     limits = c(0.7, 2.0), expand = expansion(mult = 0.01)) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0.02, 0.06))) +
  labs(x = NULL, y = "Range across life stages (°C)") +
  guides(fill = guide_legend(ncol = 1, override.aes = list(shape = 21, size = 3.2))) +
  theme_classic(base_size = 13) +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),
        axis.line = element_blank(), panel.grid = element_blank(),
        axis.text = element_text(colour = "black"),
        legend.position = c(0.005, 0.98), legend.justification = c(0, 1),
        legend.background = element_rect(fill = NA, colour = NA),
        legend.key = element_blank(), legend.key.height = unit(0.38, "cm"),
        legend.text = element_text(size = 7.5))

# (a) spans the top two grid rows; (b) and (c) share the third, 2:1 in width
panel_design <- c(area(1, 1, 2, 3),
                  area(3, 1, 3, 2),
                  area(3, 3, 3, 3))

p_species_panel <- p_facets + p_norm + p_range +
  plot_layout(design = panel_design, heights = c(1, 1, 1.25)) +
  plot_annotation(tag_levels = "a") &
  theme(plot.tag = element_text(face = "bold", size = 14))
p_species_panel

ggsave("fig/Figure2_dahlke_comparison.jpg", p_species_panel, width = 11, height = 11, dpi = 300)
ggsave("fig/Figure2_dahlke_comparison.pdf", p_species_panel, width = 11, height = 11)
```

Values behind panels (b) and (c). The immature stage is the reference stage in
panel (b), so it is assigned a value of zero without an interval.

```{r}
#| label: figure2-tables

dset_labs <- c("This study" = "This study", "Dahlke" = "Dahlke et al. 2020")

norm_tbl <- norm_avg |>
  mutate(Dataset = unname(dset_labs[as.character(Dataset)])) |>
  arrange(Dataset, LifeStage) |>
  select(Dataset, `Life stage` = LifeStage, `Species (n)` = n_sp,
         `Mean (°C)` = rel, `CI lower` = lo, `CI upper` = hi)

knitr::kable(norm_tbl, digits = 2,
             caption = "Upper thermal limit relative to the immature stage, averaged across the six species.")

range_tbl <- sp_range |>
  mutate(Species = unname(species_labs[as.character(Species)]),
         Dataset = unname(dset_labs[as.character(Dataset)])) |>
  select(Dataset, Species, `Stages compared` = n_stage, `Range (°C)` = rng) |>
  arrange(Dataset, desc(`Range (°C)`))

knitr::kable(range_tbl, digits = 2,
             caption = "Range of upper thermal limit across life stages, per species and dataset.")

knitr::kable(sp_range_avg |>
               mutate(Dataset = unname(dset_labs[as.character(Dataset)])) |>
               select(Dataset, `Species (n)` = n_sp, `Mean range (°C)` = mean_rng,
                      `CI lower` = lo, `CI upper` = hi),
             digits = 2,
             caption = "Across-species mean range and its 95% CI, per dataset.")
```

# Session information

```{r}
#| label: session-info
sessionInfo()
```
